---
title: "The Impact of Storm Events on Population Health and the Economy of the U.S."
author: "Brett Taylor"
date: "April 19, 2015"
output: pdf_document
---

  
#Synopsis
The impact of a storm may cause financial burdens, and harm to the population. Between 1996, and 2011, there were over 650,000 documented events by the federal government.  Each of the events that are documented contain how many injuries, and fatalities occurred.  They also record the cost incurred through damage to property, or to crops.  

Analysis of the events across the most top 3 most impactful event types to Population Health were Tornado's, Excessive Heat, and Floods.  The top 3 most costly event types were Floods, Hurricane/Typhoon, and Storm surges.  The review of the events shows that the majority of the storm events have occurred in the central and eastern United States, and certain locations like Oklahoma and Kansas, the Carolina's, and New Jersey and Delaware have very high levels of storm activity.    

# Data Processing   

Libraries used to process and analyze data   
```{r configure.environment,warning=FALSE,message=FALSE,echo=TRUE}
library(dplyr)
library(knitr)
library(ggplot2)
library(lubridate)
library(stats)
library(reshape2)
library(data.table)
library(ggmap)
```

This analysis utilizes 2 separate data sets.  The primary data set is the Storm Data Event data set maintained by the U.S. National Weather Service.   The secondary data set includes geographic data associated with each county in the United states, and is maintained by by United States Census Bureau.  Each of these sets of data were available in the Internet, and the data was directly downloaded and processed in the following R code chunk.  

**Obtain and read Storm Event data from the U.S. National Oceanic and Atomospheric Administration**   

```{r process.data.1,echo=TRUE,message=FALSE,cache=TRUE}
site<-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
file.name<-"StormData.csv.bz2"
data.dir <-"./data"
path.name<-paste(data.dir,file.name,sep="/")
# downloadData.R
#Step 0. Download data
## Create the ./data directory if it does not exist.
if(!file.exists(data.dir)) {
        dir.create(data.dir) 
        message(paste("Creating directory",data.dir,sep=": "))
}
## Download the data from the Internet if it does not exist.
if(!file.exists(path.name)) {
        data.source.url<-site
        download.file(url=data.source.url,destfile=path.name,method="curl")
        download.date<-date()
        unzip(path.name,exdir = data.dir)
        unzip(zipfile = path.name,list = TRUE)
         
} 
message("Reading storm data")
storm.data<-read.csv("./data/StormData.csv" )  
message("Storm data read is complete")
#read.csv reads the data appropriately, and is very slow compared to fread.  fread fails due to embeded quotes. 
storm.data<-as.data.table(storm.data)
```

**Obtain and read County geographic data from the U.S. Census Bureau**    

```{r process.data.2,echo=TRUE,message=FALSE,cache=TRUE}

#Retrieve US County data which includes the latitude, and longitude of the centroid for each county
file.name<-"Gaz_counties_national.zip"
site<-"http://www2.census.gov/geo/docs/maps-data/data/gazetteer/Gaz_counties_national.zip"
path.name<-paste(data.dir,file.name,sep="/")
counties<-read.csv("http://www2.census.gov/geo/docs/reference/codes/files/national_county.txt")
if(!file.exists(path.name)) {
        data.source.url<-site
        download.file(url=data.source.url,destfile=path.name,method="curl")
        download.date<-date()
        unzip(path.name,exdir = data.dir)
        
} 
message("Reading county data")
county.data<-read.csv(paste(data.dir,"Gaz_counties_national.txt",sep="/"),sep="\t",colClasses="character")
message("Reading county data is complete")
county.data<-as.data.table(county.data)

```
###Prepare the data
Prior to analysis, the data is transformed to allow a simple quality review.  
```{r clean.data,echo=TRUE,cache=TRUE,dependson="process.data"}

county.data$INTPTLAT <- as.numeric(county.data$INTPTLAT)
county.data$INTPTLONG<-as.numeric(county.data$INTPTLONG)
#Convert factor to date
clean.storm.data<-storm.data
clean.storm.data$BGN_DATE<-as.POSIXct(strptime(storm.data$BGN_DATE,format = "%m/%d/%Y"))
clean.storm.data$END_DATE<-as.POSIXct(strptime(storm.data$END_DATE,format = "%m/%d/%Y"))
#Create a field that contains standard state and county identifiers.
clean.storm.data$GEOID<-sprintf("%1$02.0f%2$03.0f",storm.data$STATE__,storm.data$COUNTY)
 #Narrow down the columns to what will be used for graphing.
county.data<-county.data%>%select(GEOID,INTPTLAT,INTPTLONG) 
#Combine these data.frames together based on the GEOID.  This allows map based analysis.
clean.storm.data<-left_join(clean.storm.data,county.data,by="GEOID") 
```


### Review Storm data quality
Early data in the storm event data was maintained with different levels of quality than current expectations by the National Weather Service. Through exploration of the data, it is determined that  attributes PROPDMGEXP, and CROPDMGEXP, which are the exponential cost multipliers, are good indicators of data quality changes. The valid values for PROPDMGEXP and CROPDMGEXP are "K", "M", and "B".  

```{r review.data.quality.1,echo=TRUE}


propdmg<-clean.storm.data%>%
        filter(!(PROPDMGEXP %in% c("K","M","B","")))%>%
        mutate(beg_year=year(BGN_DATE),TYPE="PROPERTY")%>%
        group_by(beg_year,TYPE)%>%
        summarize(count=n())%>%
        ungroup() 


cropdmg<-clean.storm.data%>%
        filter(!(CROPDMGEXP %in% c("K","M","B","")))%>%
        mutate(beg_year=year(BGN_DATE),TYPE="CROP")%>%
        group_by(beg_year,TYPE)%>%
        summarize(count=n())%>%
        ungroup() 
invalid.exp.data=rbind(propdmg,cropdmg)
```

Years with invalid values for Crop and Property damage exponent attributes.   

```{r review.data.quality.2,echo=TRUE}
kable(xtabs(count~.,data=invalid.exp.data))
```

The conclusion from this is that the data set better maintained after December of 1995.  This does not validate other parts of the data set, and it does give an indication of data quality. 

###Subset data based on quality metrics
The data quality review indicates that tighter controls on data quality were put in place as of calendar year 1996.  This analysis has subset-ed the data between Jan 1, 1996, and November 30, 2011.  

```{r subset.data.1,echo=TRUE,warning=FALSE}
clean.storm.data<-clean.storm.data%>%filter(BGN_DATE >= as.POSIXct("1996-01-01"))
#Map the dollar impact based on the PROPDMGEXP, and CROPDMGEXP abbreviations. 
factors.for.prop.dmg.exp<-data.frame(PROPDMGEXP=as.factor(c("K","M","B","")),prop.dmg.exp=c(1e3,1e6,1e9,0))
factors.for.crop.dmg.exp<-data.frame(CROPDMGEXP=as.factor(c("K","M","B","")),crop.dmg.exp=c(1e3,1e6,1e9,0)) 
#Join the data sets based on the abbreviations
clean.storm.data<-left_join(clean.storm.data,factors.for.prop.dmg.exp,by=c("PROPDMGEXP"="PROPDMGEXP"),copy=TRUE) 
clean.storm.data<-left_join(clean.storm.data,factors.for.crop.dmg.exp,by=c("CROPDMGEXP"="CROPDMGEXP"),copy=TRUE) 
#Compute the total property and crop damage expense
clean.storm.data<-clean.storm.data%>%
        mutate(property.damage=PROPDMG*prop.dmg.exp,
               crop.damage=CROPDMG*crop.dmg.exp)
```

The data subset is starts on date`r (min(clean.storm.data$BGN_DATE))` through `r (max(clean.storm.data$BGN_DATE))` contains `r prettyNum(nrow(clean.storm.data),big.mark=",")` observations which is sufficient to perform an accurate analysis.

# Results
The results of the analysis of the storm based events follows and focuses in 2 main areas:   

* Population Health   
* Economic Impact   

###The Consequences of storm related events on Population Health    

Storms events often create an impact on the health of the population.  Injuries, and fatalites are often the outcome of an event.  The number of injuries and fatalities are tracked by the National Weather Service, and the top ten event types are displayed below.

```{r pop_health_data,echo=TRUE,message=FALSE,warning=FALSE}

by.env <-clean.storm.data%>%group_by(EVTYPE)%>%
        summarize(Injury=sum(INJURIES),Fatality=sum(FATALITIES))%>%
        ungroup()%>%
        arrange(desc(Fatality+Injury))
top.10<-head(by.env,10)
#Sequence the top 10 by Event Type - largest impact to smallest
top.10$EVTYPE<-factor(top.10$EVTYPE,levels=unique(top.10$EVTYPE))
top.10.injuries.fatalities<-melt(top.10,id=c("EVTYPE"),impact=c("INJURIES","FATALITIES"))
 
```

Below is the plot of the impact of storm events on the U.S. Population's Health.    

```{r pop_health,echo=TRUE,message=FALSE,warning=FALSE}

p1 <-ggplot(top.10.injuries.fatalities,aes(x=EVTYPE,y=value,fill=variable))+
        geom_bar(stat="identity")+
        theme_bw()+
        labs(x="Storm Event Type",y="Count",
             title="Population Health Impact of Storm Events\n USA Top 10 Event Types\n1996 through 2011",             
             fill="Impact")+
        theme(axis.text.x=element_text(angle=30, hjust=1,vjust=1,size=8),
              legend.position=c(1,1),legend.justification=c(1,1)) 
 
print(p1)
```

**Figure 1** The impact to USA population health caused by the top 10 storm event types   

The population health of the United States is most impacted by `r top.10$EVTYPE[1]`.  The topy 3 most sigificant event types that impact population health are Tornados, Excessive Heat and Floods.  

*******
###Economic consequences of storm related events
To determine the economic consequences of storm events, we need to understand the cost of property damage, and the cost of crop damage. 

*Each injury and fatality has an economic impact on the society. These data are not included in this analyis, and will not be considered at this time.  If you would like to review potential impact of [safety](http://safety.fhwa.dot.gov/facts_stats/t75702.cfm) this website is a good starting place. *


```{r cost_impact_data,echo=TRUE,message=FALSE,warning=FALSE}


cost.by.env <-clean.storm.data%>%
        group_by(EVTYPE)%>%
        summarize( 
                  tot.property.damage=sum(property.damage),
                  tot.crop.damage=sum(crop.damage))%>%
        ungroup()%>%
        arrange(desc(tot.property.damage+ tot.crop.damage))
top.10.cost<-head(cost.by.env,10)
#Sequence the top 10 by Event Type - largest cost impact to smallest
top.10.cost$EVTYPE<-factor(top.10.cost$EVTYPE,levels=unique(top.10.cost$EVTYPE))

top.10.cost<-melt(top.10.cost,id=c("EVTYPE"))
```

Below is the plot of the economic impact of storm events.    
```{r cost_impact,echo=TRUE,message=FALSE,warning=FALSE}
p2 <-ggplot(top.10.cost,aes(x=EVTYPE,y=value/1e9,fill=variable))+
        geom_bar(stat="identity",color="black")+
        theme_bw()+
        labs(x="Storm Event Type",y="Cost in $Billion",
             title="Economic Impact of Storm Events\n USA Top 10 Event Types\n1996 through 2011",             
             fill="Impact")+
        theme(axis.text.x=element_text(angle=30, hjust=1,vjust=1,size=8),
              legend.position=c(1,1),legend.justification=c(1,1))  
 
print(p2)

```

**Figure 2** The impact fiscal impact of storm events in the USA   

The top three storm event types that create the largest cost impact are Floods, Hurricanes & Typhoons, and Storm Surges.  The major impact in this analysis is due to property damage.

###Distribution of Storm Events across the continental United States
Not all areas in the United States suffer from the same level of storm events.   The distribution of storm events is demonstrated through a geographical mapping of *all* storm events between 1996, and 2011.  While this does not indicate the impact of the event, it does present the frequency of events of all types.   

```{r across_usa,echo=TRUE,message=FALSE,warning=FALSE,fig.width=10,fig.height=10}

#test.data<-clean.storm.data[sample(1:nrow(clean.storm.data),200000,replace = FALSE),]
test.data<-as.data.table(clean.storm.data)
USA.map<-qmap("United States",zoom=4,color="bw",
     base_layer = ggplot(aes(x=INTPTLONG, y=INTPTLAT), data = test.data))
USA.map + 
        stat_density2d(aes(x = INTPTLONG, y = INTPTLAT, fill = ..level.., alpha = ..level..),
                       geom="polygon",
                       data = test.data) +
        scale_fill_gradient(low = "green", high = "red")+  
        theme(legend.position = "none", axis.title = element_blank(), text = element_text(size = 12))
```

**Figure 3** The distribution of all storm events across the continental United States

The top areas with the most significant number of storm events are in the central and north eastern United States.  The western United States demonstrates small areas of storm events during the time period reviewed.  So if you want to live in a safe, and beautiful location, join us in **Idaho**!
